FAIR fission track analysis with geochron@home

### 2025-08-21 geochron@home/manuscript/paper.tex
Side caption for Table 1 in g@h paper

geochron@home serves four different purposes. It can be used (1) to
count fission tracks in `private mode', i.e. hidden from other users
on the internet; (2) to archive fission track images and counts for
inspection by other users; (3) to create tutorials for new students of
the fission track method; and (4) to generate crowd-sourced fission
track data, by serving randomly selected selections of images to
citizen scientists. We illustrate these four applications with
examples that demonstrate (1) geochron@home's ability to compare and
combine fission track counts for multiple users within a lab group;
(2) the value of the geochron@home archive in the peer review system;
(3) the use of simple tutorials in teaching novice users how to count
fission tracks; and (4) the `wisdom of crowds' in fission track
ratio estimation.\medskip

@@

Unlike most other geochronometers, which require mass spectrometers to
estimate parent-daughter ratios, fission tracks are observed under an
optical microscope and counted by a human observer.  In recent years,
digital microscopy has moved fission track data acquisition from the
objective lens of a microscope to the computer screen
\citep{gleadow2009, vanranst2019, gleadow2019}. Ongoing developments
in artificial intelligence generate further opportunities to improve
the throughput and accuracy of fission track data
\citep{nachtergaele2020}. But despite the richness of the digital
datasets produced by these novel tools, fission track data are still
reported as small summary tables. These tables require an unnecessary
degree of trust between the `producer' and `consumer' of the
data.\medskip

@@

  In addition to the sequence of \texttt{.jpeg} images, a low level
  `grain' folder can also include an optional file called
  \texttt{roi.json} containing the vertices of a default region of
  interest for spontaneous (and/or induced) track counting. The
  database structure can also accommodate existing fission track
  counting results.  For example, if a user has already counted their
  fission tracks in Fission Track Studio, then they can store those
  results in a \texttt{.json} file at the `sample' level
  directory. Because Fission Track Studio stores its results in an
  \texttt{.xml} format, a second conversion script is needed to
  translate those results into an equivalent \texttt{.json} format.

@@

\begingroup
\captionof{table}{Summary table of the crowd sourcing results for the
  different users (rows) and grains (columns).  The last row is the
  ratio of the median count to PV's count for each grain.  The last
  column is the ratio of the pooled count over grains compared with PV
  for the same grains.}
\label{tab:crowdtable}
\renewcommand{\arraystretch}{0.7}
\begin{scriptsize}
  \input{../output/crowdtable.txt}
\end{scriptsize}
\endgroup

@@

However, in spite of the under-counting by the students, the
\emph{ratios} of the track counts between different grains are
generally quite close to PV's ratios, especially when comparing grains
with similar track densities.  This is apparent from
Table~\ref{tab:crowdtable} (looking along columns).  In grains
containing few tracks (left hand side of the table), the median of the
student counts is within 20\% of PV's counts.  However, this gradually
levels off to 50\% at higher track densities (right hand side of the
table). Figure~\ref{fig:23vs25}b shows a scatter plot of pairs of
counts for two grains (23 and 25) made by 36 students who counted both
of them. The two grains were chosen because they have similar track
counts for both PV and the median student. They both have a median of
28 tracks per grain compared with PV's pair of 52 and 50, which is
highlighted in blue. There is a positive correlation between the
pairs, consistent with systematic observer effects (i.e., people who
count a low value in one grain tend to count similarly low in the
other, and vice versa).  Comparing other pairs of grains shows similar
results, with correlations varying between 0.4 and 0.9. Pairs of
grains with similar track densities are more strongly correlated than
those with dissimilar track densities. In all cases we have looked at,
the residual variation, after allowing for systematic differences
between students, is consistent with Poisson variation.\medskip

As mentioned in Section~\ref{sec:architecture}, geochron@home stores
the actual track positions marked by the users. This raw data can be
downloaded as a \texttt{.json} file and inspected in detail.
Figure~\ref{fig:23vs25}b) shows two-dimensional histograms for the x-y
positions of all the track positions generated by the students in
grains~23 and 25. Visual comparison of the histogram with the optical
image confirms that the students unanimously identified the most
obvious semitracks, which contain a clearly visible etch pit and
tail. Shorter and fainter tracks received fewer clicks. Datasets like
this can be used to replace integer counts of fission tracks with
probabilities, reflecting the ambiguity of some fission track
datasets. Figure~\ref{fig:23vs25} also shows that some students
counted the tails of the fission tracks rather than their etch pits,
despite being told the opposite in the tutorial. Fixing this issue
will require some improvements to the tutorial pages
(Section~\ref{sec:outlook}).\medskip

@@

\begin{enumerate}
\item Length measurements. The geochron@home archive already
  contains horizontally confined fission track measurements
  \citep{tamer2025}.  However, these results must be generated
  externally (e.g., using Fission Track Studio) and uploaded via a
  \texttt{.json} file.  The virtual microscope curently lacks the
  functionality to generate length data within
  geochron@home. This functionality will be added in a future
  update.
\item Dpar and Dper. Etch pits are currently stored as simple sets of
  x- and y-coordinates.  In reality, etch pits have a finite length
  (`Dpar') and width (`Dper'), which serve as useful indicators for
  the horizontal etch rates along the c-axis and parallel to it
  \citep{donelick1993}. Functions will be added to measure and
  visualise this type of data in geochron@home.
\item Interactive tutorial pages and embedded quality
  control. Section~\ref{sec:crowdsourcing} shows that the
  crowd-sourcing idea is not yet ready for general use. Two
  improvements will be made to make its results more reliable. First,
  a `quiz' to the tutorial pages to ensure that novice users do not
  count the tails but the etch pits of fission tracks. Only users who
  click on the `correct' features in an unlabelled set of images will
  be allowed to count new samples. Second, future citizen scientists
  will occasionally be served reference images of known track
  density. Their counts for these `test' images will be used to
  identify trustworthy analysts (top of Table~\ref{tab:crowdtable}),
  and flag unreliable ones (bottom of Table~\ref{tab:crowdtable}).
\item Machine learning. Data science is experiencing an artificial
  intelligence revolution that has already started to transform the
  fission track method \citep{nachtergaele2020}.  Convolutional neural
  networks must be trained with example data. geochron@home
  is ideally suited for this task. Section~\ref{sec:crowdsourcing}
  showed how the collective wisdom of multiple fission track analysts
  can label fission track images with probabilities rather than
  counts. This data format is close to the form in which data are
  treated within an AI algorithm.
\item Inverse counting. Once trained on historical data, AI algorithms
  can be used to count fission tracks automatically. Following the
  model of Fission Track Studio \citep{gleadow2009, gleadow2019},
  machine learning can be used to reverse the fission track counting
  process. Instead of asking users to count the fission tracks in a
  sample, the software can ask them to check the results proposed by
  an AI algorithm, and to remove any features that are \emph{not}
  fission tracks.  Regardless of whether fission tracks were counted
  manually or with a machine, the value of the geochron@home archive
  remains the same. It is important to document data so that samples
  can be reanalysed in the future, for example when a new and improved
  generation of machine learning algorithms becomes available.
\end{enumerate}

@@

This paper introduced geochron@home, a software platform for FAIR
fission track analysis. We demonstrated four different applications of
this platform using real data. Putting the FAIR data paradigm into
practice, all the imagery, counts and source code for this paper are
publicly available. Using these resources, the reader can reproduce
all the results that were presented in this publication.  We encourage
other geochronologists to follow this example.  FAIR data promises to
address the reproducibility crisis in science
\citep{miyakawa2020}.\medskip

@@

\codedataavailability{geochron@home is free software released under
  the GPL-3 license. The package and its source code are available
  from \url{https://github.com/pvermees/geochron-at-home} (last
  access: July~21, 2025).  The raw data (imagery) are available at the
  geochron@home archive (\url{https://github.com/pvermees/GaHa}, last
  access: July~21, 2025).  R-scripts to reproduce the figures are
  provided in the supplementary information
  (\url{https://github.com/pvermees/supplements}, last access:
  July~21, 2025).}

@@

\authorcontribution{PV designed the study, acquired the funding and
  counted fission tracks. JH created geochron@home. TB expanded
  geochron@home and wrote the accompanying microscope image
  acquisition software. RG derived Equation~\ref{eq:NO} and verified
  the other calculations. AC provided the samples, prepared the
  training data and counted fission tracks. PV and RG wrote the paper
  with input from the other authors.}

>>>>

geochron@home serves four different purposes. It can be used (1) to
count fission tracks in `private mode', i.e. hidden from other users
on the internet; (2) to archive fission track images and counts for
inspection by other users; (3) to create tutorials for new students of
the fission track method; and (4) to serve randomly selected
selections of images to citizen scientists. We illustrate these four
applications with examples that demonstrate (1) geochron@home's
ability to compare and combine fission track counts for multiple users
within a lab group; (2) the value of the geochron@home archive in the
peer review system; (3) the use of simple tutorials in teaching novice
users how to count fission tracks; and (4) the opportunities and
challenges of crowd-sourced fission track analysis.\medskip

@@

Unlike most other geochronometers, which require mass spectrometers to
estimate parent-daughter ratios, fission tracks are observed under an
optical microscope and counted by a human observer.  In recent years,
digital microscopy has moved fission track data acquisition from the
ocular lenses of a microscope to the computer screen
\citep{gleadow2009, vanranst2019, gleadow2019}. Ongoing developments
in artificial intelligence generate further opportunities to improve
the throughput and accuracy of fission track data
\citep{nachtergaele2020}. But despite the richness of the digital
datasets produced by these novel tools, fission track data are still
reported as small summary tables. These tables require an unnecessary
degree of trust between the `producer' and `consumer' of the
data.\medskip

@@

  In addition to the sequence of \texttt{.jpeg} images, a low level
  `grain' folder can also include an optional file called
  \texttt{roi.json} containing the vertices of a default region of
  interest for spontaneous (and/or induced) track counting. The
  database structure can also accommodate existing fission track
  counting results.  For example, if a user has already counted their
  fission tracks in Fission Track Studio, then they can store those
  results in a \texttt{.json} file at the `sample' level
  directory. Because Fission Track Studio stores its results in an
  \texttt{.xml} format, a second conversion script was created to
  translate those results into an equivalent \texttt{.json} format.

@@

@@

However, in spite of the under-counting by the students, the
\emph{ratios} of the track counts between different grains are often
close to PV's ratios, especially when comparing grains with similar
track densities.  This is apparent from Table~\ref{tab:crowdtable}
(looking along columns).  In grains containing few tracks (left hand
side of the table), the median of the student counts is within 20\% of
PV's counts.  However, this gradually levels off to 50\% at higher
track densities (right hand side of the
table). Figure~\ref{fig:23vs25}a shows a scatter plot of pairs of
counts for two grains (23 and 25) made by 36 students who counted both
of them. The two grains were chosen because they have similar track
counts for both PV and the median student. They both have a median of
28 tracks per grain compared with PV's pair of 52 and 50, which is
highlighted in blue. There is a positive correlation between the
pairs, consistent with systematic observer effects (i.e., people who
count a low value in one grain tend to count similarly low in the
other, and vice versa).  Comparing other pairs of grains shows similar
results, with correlations varying between 0.4 and 0.9. In all cases
we have looked at, the residual variation, after allowing for
systematic differences between students, is consistent with Poisson
variation.\medskip

As mentioned in Section~\ref{sec:architecture}, geochron@home stores
the actual track positions marked by the users. This raw data can be
downloaded as a \texttt{.json} file and inspected in detail.
Figure~\ref{fig:23vs25}b shows two-dimensional histograms for the x-y
positions of all the track positions generated by the students in
grains~23 and 25. Visual comparison of the histograms with the optical
images confirms that the students unanimously identified the most
obvious semitracks, which contain a clearly visible etch pit and
tail. Shorter and fainter tracks received fewer clicks. Datasets like
this can be used to replace integer counts of fission tracks with
probabilities, reflecting the ambiguity of some fission track
datasets. Figure~\ref{fig:23vs25} also shows that some students
counted the tails of the fission tracks rather than their etch pits,
despite being told the opposite in the tutorial. Fixing this issue
will require some improvements to the tutorial pages
(Section~\ref{sec:outlook}).\medskip

\noindent\begin{minipage}{.65\linewidth}
\renewcommand{\arraystretch}{0.7}
\begin{scriptsize}
  \input{../output/crowdtable.txt}
\end{scriptsize}
\end{minipage}
\begin{minipage}{.35\linewidth}
\captionof{table}{Summary table of the crowd sourcing results for the
  different users (rows) and grains (columns).  The last row is the
  ratio of the median count to PV's count for each grain.  The last
  column is the ratio of the pooled count over grains compared with PV
  for the same grains.}
\label{tab:crowdtable}
\end{minipage}

@@

\begin{enumerate}
\item Length measurements. The geochron@home archive already
  contains horizontally confined fission track measurements
  \citep{tamer2025}.  However, these results must be generated
  externally (e.g., using Fission Track Studio) and uploaded via a
  \texttt{.json} file.  The virtual microscope curently lacks the
  functionality to generate length data within
  geochron@home. This functionality will be added in a future
  update.
\item Dpar and Dper. Etch pits are currently stored as simple sets of
  x- and y-coordinates.  In reality, etch pits have a finite length
  (`Dpar') and width (`Dper'), which serve as useful indicators for
  the horizontal etch rates along the c-axis and parallel to it
  \citep{donelick1993}. Functions will be added to measure and
  visualise this type of data in geochron@home.
\item Interactive tutorial pages and embedded quality
  control. Section~\ref{sec:crowdsourcing} shows that the
  crowd-sourcing idea is not yet ready for general use. Two
  improvements will be made to make its results more reliable. First,
  a `quiz' will be added to the tutorial pages to ensure that novice
  users do not count the tails but the etch pits of fission
  tracks. Only users who click on the `correct' features in an
  unlabelled set of images will be allowed to count new
  samples. Second, future citizen scientists will occasionally be
  served reference images of known track density. Their counts for
  these `test' images will be used to identify trustworthy analysts
  (top of Table~\ref{tab:crowdtable}) and flag unreliable ones (bottom
  of Table~\ref{tab:crowdtable}).
\item Machine learning. Data science is experiencing an artificial
  intelligence (AI) revolution that has already started to transform
  the fission track method \citep{nachtergaele2020}.  Convolutional
  neural networks must be trained with example data. geochron@home is
  ideally suited for this task. Section~\ref{sec:crowdsourcing} showed
  how the opinion of multiple fission track analysts can be combined
  to label fission track images with probabilities rather than
  counts. This data format is close to the form in which data are
  treated within an AI algorithm.
\item Inverse counting. Once trained on historical data, AI algorithms
  can be used to count fission tracks automatically. Following the
  model of Fission Track Studio \citep{gleadow2009, gleadow2019},
  machine learning can be used to reverse the fission track counting
  process. Instead of asking users to count the fission tracks in a
  sample, the software can ask them to check the results proposed by
  an AI algorithm, and to remove any features that are \emph{not}
  fission tracks.  Regardless of whether fission tracks were counted
  manually or with a machine, the value of the geochron@home archive
  remains the same. It is important to document data so that samples
  can be reanalysed in the future, for example when a new and improved
  generation of machine learning algorithms becomes available.
\end{enumerate}

@@

This paper introduced geochron@home, a software platform for FAIR
fission track analysis. We demonstrated four different applications of
this platform using real data. Putting the FAIR data paradigm into
practice, all the imagery, counts and source code for this paper are
publicly available on the geochron@home archive. Using these
resources, the reader can reproduce all the results that were
presented in this publication.  We encourage other geochronologists to
follow this example.  FAIR workflows promise to address the
reproducibility crisis in science \citep{miyakawa2020}.\medskip

@@

\codedataavailability{geochron@home is free software released under
  the GPL-3 license. The package and its source code are available
  from \url{https://github.com/pvermees/geochron-at-home} (last
  access: August~21, 2025).  The raw data (imagery) are available at
  the geochron@home archive (\url{https://github.com/pvermees/GaHa},
  last access: August~21, 2025).  R-scripts to reproduce the figures
  are provided in the supplementary information
  (\url{https://github.com/pvermees/supplements}, last access:
  August~21, 2025).}

@@

\authorcontribution{PV designed the study, acquired the funding and
  counted fission tracks. JH created geochron@home. TB expanded
  geochron@home and wrote the accompanying microscope image
  acquisition software. RG derived Equation~\ref{eq:NO}, designed
  Table~\ref{tab:crowdtable} and verified the other calculations. AC
  provided the samples, prepared the training data and counted fission
  tracks. PV and RG wrote the paper with input from the other
  authors.}
